---
title: "Project 2"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

This is the dataset you will be working with:
```{r message = FALSE}
olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')

olympic_gymnasts <- olympics %>% 
  filter(!is.na(age)) %>%             # only keep athletes with known age
  filter(sport == "Gymnastics") %>%   # keep only gymnasts
  mutate(
    medalist = case_when(             # add column for success in medaling
      is.na(medal) ~ FALSE,           # NA values go to FALSE
      !is.na(medal) ~ TRUE            # non-NA values (Gold, Silver, Bronze) go to TRUE
    )
  )

```

**Question:** Are there age differences for male and female Olympic gymnasts who were successful or not in earning a medal, and how has the age distribution changed over the years? 



**Approach:** First we will show the distribution of age of the athletes versus the sex of them in a violin plot (```geom_violin()```) and we will use status of medal received or not as a ```color``` feature in the plot. Using a violin plot will help us to visualize many distributions at once and it gives an overall shifts among the distributions.

For the second part, we will use faceted boxplot to show the distribution of age of the athletes over the years of the Games and using a label of whether medal received or not and the sex of the athlete. This partitions the whole plot into a matrix of panels with a subset of the data and makes it easier to get detailed overall shift information among the distributions.

**Analysis:**

```{r }
# violin plot showing age distribution for male and female athletes.
ggplot(olympic_gymnasts,aes(sex,age,fill=medalist))+
  geom_violin()+
  scale_x_discrete(name='Sex of the athlete',
                   labels= c('F','M')
                   )+
  scale_y_continuous(name = 'Age of the athlete (years)'
                    )+
    ggtitle('Age distribution for male and female athletes \n and their medal status')+
  theme_bw(12)

                   


```

```{r fig.width=10,fig.height=11}
#Age distribution over the years
ggplot(olympic_gymnasts,aes(year,age,group=year))+
  geom_boxplot(alpha=0.75)+
  facet_wrap(~medalist+sex,scales='free',
labeller = as_labeller(c('TRUE' = 'Medaled', 'FALSE' = 'Did not medal','M'='Male','F'='Female')))+
  xlab('year of the Games')+
 ylab('age of athletes (in years)')+
ggtitle('Age distribution of athletes over the years and their medal win status')+
  theme_bw(12)


               

```


