---
title: "Project 2"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

This is the dataset you will be working with:
```{r message = FALSE}
olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')

olympic_gymnasts <- olympics %>% 
  filter(!is.na(age)) %>%             # only keep athletes with known age
  filter(sport == "Gymnastics") %>%   # keep only gymnasts
  mutate(
    medalist = case_when(             # add column for success in medaling
      is.na(medal) ~ FALSE,           # NA values go to FALSE
      !is.na(medal) ~ TRUE            # non-NA values (Gold, Silver, Bronze) go to TRUE
    )
  )

```

**Question:** Are there age differences for male and female Olympic gymnasts who were successful or not in earning a medal, and how has the age distribution changed over the years? 


**Introduction:** We are working with the ```olympics``` dataset, which provides Olympic Game records of athletes and medal results for all Games from Athen 1896 to Rio 2016. For our purpose, we have created a dataset named ```olympic_gymnasts``` from ```olympics``` and it contains 25,528 records of only gymnasts with known age and we added a new column named ```medalist``` that indicates success in medaling at the end of the table. In this dataset, each row corresponds to one athlete, and there are 16 columns providing information about the athlete, the Games and the events. Data on athlete include their unique ID, name, sex, age, height in cm, weight in kg, team they represent, and the National Olympic Committee region they come from. Additional information on the events each athlete participated in, included sport name, event name, and medal status and information on each Game includes name of the game, year it was held, season, and host city.

To answer the first question we will be working with variables: sex (given as M or F), age (given in numeric value) and whether the athlete received a medal or not (column name ```medalist```) (given as TRUE or FALSE). For the second question we will use year of the Games, age of athletes, sex, and ```medalist``` columns. 

**Approach:** First we will show the distribution of age of the athletes versus the sex of them in a violin plot (```geom_violin()```) and we will use status of medal received or not as a ```color``` feature in the plot. Using a violin plot will help us to visualize many distributions at once and it gives an overall shifts among the distributions.

For the second part, we will use faceted boxplot to show the distribution of age of the athletes over the years of the Games and using a label of whether medal received or not and the sex of the athlete. This partitions the whole plot into a matrix of panels with a subset of the data and makes it easier to get detailed overall shift information among the distributions.

**Analysis:**

```{r }
# violin plot showing age distribution for male and female athletes.
ggplot(olympic_gymnasts,aes(sex,age,fill=medalist))+
  geom_violin()+
  scale_x_discrete(name='Sex of the athlete',
                   labels= c('F','M')
                   )+
  scale_y_continuous(name = 'Age of the athlete (years)'
                    )+
    ggtitle('Age distribution for male and female athletes \n and their medal status')+
  theme_bw(12)

                   


```

```{r fig.width=10,fig.height=11}
#Age distribution over the years
ggplot(olympic_gymnasts,aes(year,age,group=year))+
  geom_boxplot(alpha=0.75)+
  facet_wrap(~medalist+sex,scales='free',
labeller = as_labeller(c('TRUE' = 'Medaled', 'FALSE' = 'Did not medal','M'='Male','F'='Female')))+
  xlab('year of the Games')+
 ylab('age of athletes (in years)')+
ggtitle('Age distribution of athletes over the years and their medal win status')+
  theme_bw(12)


               

```

**Discussion:** From the violin plot it is evident that age distribution for male and female gymnasts are different for their medal winning/non-winning status. We see more female gymnasts in the age group of 15-20 who won/did not win the medal and the male gymnasts are mainly in the age group of 20-25. This means, we see more younger female gymnasts in the Games and it can be attributed to various factors such as female gymnasts pick up the game very early compared to males and reach at their peak physical condition much earlier. Interestingly, we do not see any difference in age distribution on the medal winning status for female gymnasts (both pink and blue violins show similar distribution for female). This is the same case for male gymnasts too. 

In the faceted boxplots we see age distribution of male and female gymnasts over the years and their medal winning status. For male gymnasts, the age range until year 1940 was 20-30 for medal winners, between year 1950-1975 the age range increased slightly to 22-32, and then we see a dip between year 1975-2016 with a range of 20-26 years of age.We see a similar pattern for male non-medaled gymnasts. For female gymnasts who won medals, the age range was 14-24 around year 1925, between year 1950-1962 the age range was 20-28 and then between 1965-2016 the age range saw a dip to the range of 15-22. We see a similar pattern for female non-medaled gymnasts. 

Hence, we can conclude that ages for male and female gymnasts were a little higher for earlier Olympic Games, especially right after the World War I and II, which could be attributed to the fact that most young people went to wars and took time to recover and were unable to participate immediately after that. In recent years (after 1975) more young gymnasts (difference in age ~5-7 years within each sex) are representing their teams and it can be an outcome of factors like better coaching at young age or increased interest among young generation or social media impact of Olympic Games related news. 
